Factors affecting inferences on natural mortality and associated environmental effects

### 2023-10-04 Ecov_study/mortality/paper/paper.Rmd
saving work on M-ecov working paper

The covariate effect is one of 3 alternative values in the operating models:
\begin{itemize}
\item $\beta_\text{Ecov} \in \{0,0.25,0.5\}$
\end{itemize}

@@

We calculated median relative errors of
\begin{itemize}
\item $\beta_\text{Ecov}$, the effect of environmental covariates on natural mortality, 
\item $\beta_M$, the mean log-natural mortality rate,
\item annual natural mortality rate,
\item annual spawning stock biomass, and
\item annual fully-selected fishing mortality rate.
\end{itemize}
For the $i$th simulated data the relative error for a parameter $\theta$ provided from the fitted estimation model is
$$
{\text{RE}_i}\left(\theta\right) = \frac{\widehat \theta_i - \theta_i}{\theta_i}
$$
and measured bias as the median relative error and constructed 95\% confidence intervals using the binomial distribution approach as in \citet{stockmiller21}. If the confidence interval contains zero we do not have evidence of bias from the simulation study. We used estimates from all simulations that satisfied the first type of convergence described above. We were not concerned about more restrictive types of convergence because some estimation models would not be expected to satisfy these criteria because of mis-specified structures. For example, an estimation model that includes survival random effects fitted to a set of observations simulated without survival random effects should have trouble estimating variance of the survival random effects, but the estimates of the parameters and derived output described above, might be reliable.

@@

@@

@@

@@

@@

>>>>

The covariate effect is one of 3 alternative values in the operating models:
\begin{itemize}
\item $\beta_\text{Ecov} \in \{0,0.25,0.5\}$
\end{itemize}
The parameters defining the simulated covariate time series, size of the covariate effect, and any natural mortality random effects result in a range of different levels of variation in annual natural mortality rates (Figure \ref{M_example}).

@@

We calculated median relative errors of
\begin{itemize}
\item $\beta_\text{Ecov}$, the effect of environmental covariates on natural mortality, 
\item $\beta_M$, the mean log-natural mortality rate,
\item annual natural mortality rate,
\item annual spawning stock biomass, and
\item annual fully-selected fishing mortality rate.
\end{itemize}
Results for fishing mortality rate are provided in the Supplementary Materials. For the $i$th simulated data the relative error for a parameter $\theta$ provided from the fitted estimation model is
$$
{\text{RE}_i}\left(\theta\right) = \frac{\widehat \theta_i - \theta_i}{\theta_i}
$$
and measured bias as the median relative error and constructed 95\% confidence intervals using the binomial distribution approach as in \citet{stockmiller21}. If the confidence interval contains zero we do not have evidence of bias from the simulation study. We used estimates from all simulations that satisfied the first type of convergence described above. We were not concerned about more restrictive types of convergence because some estimation models would not be expected to satisfy these criteria because of mis-specified structures. For example, an estimation model that includes survival random effects fitted to a set of observations simulated without survival random effects should have trouble estimating variance of the survival random effects, but the estimates of the parameters and derived output described above, might be reliable.

For natural mortality rate, SSB, and fully-selected fishing mortality rate, we summarized results for each of the annual values, but we present results only for estimates from the first year (start), year 21 (middel), and the last year (end), because there were no appreciable differences between the results for these three years and those for the other years.

@@

We found high accuracy for estimation of the mean natural mortality rate parameter ($\beta_M$) for all EM process errors assumptions when OMs had step changes in fishing mortality, lower uncertainty in index and age composition observations, and either R or R+M process errors (Figure \ref{mean_M_bias}). The most variation in estimates occurred when fishing and Higher uncertainty in index and age composition observations. For OMs with R+S process errors, the most reliable estimation of $\beta_M$ was obtained when the EM also assumed R+S process errors across all other factors. For OMs with R+M process errors, the matching assumption for the EM only showed the best reliability when there was low uncertainty in index and age composition observations and fishing mortality was constant.

@@

We present results for error in annual natural mortality rate conditional on three alternative EM configurations for the natural mortality parameters: 1) mean natural mortality rate parameter is fixed at the true value ($\beta_M = log(0.2)$) and no covariate effect is assmed ($\beta_\text{Ecov} = 0$),  2) $\beta_M = log(0.2)$ and $\beta_\text{Ecov}$ is estimated, and 3) both $\beta_M$ and $\beta_\text{Ecov}$ are estimated.

When OMs and EMs assume $\beta_M = log(0.2)$ and $\beta_\text{Ecov} = 0$, there is no annual variation in natural mortality for OMs with R or R+S process errors simulated or assumed in the EM and, therefore, estimation bias is not possible (Figure \ref{annual_M_bias_M_fixed_beta_fixed}). We also observe little or no evidence of bias (confidence intervals include 0) for R or R+S EMs for any OMs even when the OMs included an effect of the covariate on natural mortality ($\beta_\text{Ecov} > 0$). Including process error on M (R+M) produces more variability in errors for R and R+S EMs than including increased level of effect of the covariate on M. For R and R+S OMs, errors in annual M for R+S EMs were less variable than those for R EMs.

R+M EMs fit to R OMs with constant fishing mortality, or a step change in fishing mortality and lower uncertainty in index and age composition observations exhibited no bias in annual estimation of M indicating that the variance of the estimated random effects for M in these EMs collapsed to 0. R+M EMs were the only EMs to exhibit differences in the sign of the median errors across the time series. When OMs had a step change in fishing mortality and higher uncertainty in index and age composition observations, we observed positive median errrors at the beginning of the time series and lower median errors in later years. However, whether there was evidence of bias, depended on the uncertainty in covariate observations and the degree of variability in the latent covariate. We observed negative median relative errors in annual M estimates for R+M EMs fit to R+S OMs with lower uncertainty in index and age composition observations across all levels of simulated effects of the covariate, but evidence of bias was strongest at the beginning of the time series.

When $\beta_\text{Ecov}$ was estimated, R EMs performed worse for R+S OMs with evidence of negative bias for OMs with cosntant fishing mortality, lower uncertainty in index and age composition observations, the largest covariate effect size, lower variability in the latent covariate, and larger uncertainty in the covariate observations (Figure \ref{annual_M_bias_M_fixed_beta_estimated}). There was little difference in the results for R+S EMs whether $\beta_\text{Ecov}$ was estimated or not. The results for R+M EMs were generally similar to those when $\beta_\text{Ecov}=0$ was assumed, except that the differences between the estimates at the beginning of the time series and later years for certain OM configurations did not occur.

Allowing the EMs to estimate $\beta_M$ resulted in very large variability in estimates for all EMs for R and R+M OMs with constant fishing mortality and higher uncertainty in index and age composition observations (Figure \ref{annual_M_bias_M_estimated_beta_estimated}). The same varibilities occurred for R+S OMs, but strong bias was estimated for R and R+M EMs. Annual M estimation was most reliable when OMs had step changes in fishing mortality and lower uncertainty in index and age composition observations. For R and R+M OMs with those characteristics, all EMs generally provided accurate estimation of annual M, but only R+S EMs provided accurate estimation for R+S OMs. Little or no bias in annual M estimation was observed for R+S EMs across all OM process error assumptions as long as there was a step change in fishing mortality and lower uncertainty in indices and age composition.

@@

Like natural mortality, we present results for error in spawning biomass conditional on three alternative EM configurations for the natural mortality parameters: 1) mean natural mortality rate parameter is fixed at the true value ($\beta_M = log(0.2)$) and no covariate effect is assmed ($\beta_\text{Ecov} = 0$),  2) $\beta_M = log(0.2)$ and $\beta_\text{Ecov}$ is estimated, and 3) both $\beta_M$ and $\beta_\text{Ecov}$ are estimated.

When EMs assume $\beta_M = log(0.2)$ and $\beta_\text{Ecov} = 0$, there is little evidence of bias for R and R+M OMs except when there is a step change in fishing mortality and higher uncertainty in index and age composition observations, particularly at the end of the time series (Figure \ref{SSB_bias_M_fixed_beta_fixed}). We observed a similar trend in median relative error for R+S OMs, but there was more evidence of positive bias at the beginning of the time series whereas the confidence intervals for the negative median errors at the end of the time series often included 0. However, there was more indication of positive bias of the incorrect process error assumption of the EMs for the R+S OMs. For R+S EMs fit to R+S OMs, there was indication of small positive bias at the beginning of the time series when there was constant fishing mortality and higher uncertainty in index and age composition observations.

When EMs assume $\beta_M = log(0.2)$, but estimate $\beta_\text{Ecov}$, the median errors for SSB for R and and R+M OMs are similar to those when $\beta_\text{Ecov} = 0$ (Figure \ref{SSB_bias_M_fixed_beta_estimated}). For R+S OMs, the median errors for EMs that also assume R+S process errors are also similar to those with no covariate effect assumed. However, for R and R+M EMs fit to R+S OMs with constant fishing mortality rate and lower uncertainty in indices and age composition observations, we observed evidences of negative bias when higher covariate observation uncertainty and lower variation in the latent covariate was simulated and positive bias for other configurations of the covariate and corresponding observation uncertainty. When R+S OMs had constant fishing mortality rate and higher uncertainty in indices and age composition observations or step changes in fishing mortality and lower uncertainty in indices and age composition, R and R+M EMs often provided positively biased SSB estimates when there was lower uncertainty in covariate observations.

When EMs estimated both $\beta_M$ and $\beta_\text{Ecov}$, we observed large variation in the errors in SSB under the same OM configurations where we observed large variation in errors for annual natural mortality rates (Figure \ref{SSB_bias_M_estimated_beta_estimated}). Similarly, we observed little or no bias in SSB estimation for R+S EMs across all OM process error assumptions as long as there was a step change in fishing mortality and lower uncertainty in indices and age composition.

@@

The lack of bias observed for annual M, using R+S EMs when no effect was assumed must be due to the annual M being equal to the assumed value (0.2) on average across simulations because the simulated environmental covariate has mean 0. However, if one were to condition on the covariate time series we would expect biased estimation of annual M when there was a covariate effect.

@@

\begin{landscape}
\begin{figure}
\caption{Example simulations of annual natural mortality rates that may be a function of a temporally varying environmental covariate and autoregressive random effects.}\label{M_example}
\begin{center}
\includegraphics[width = \textwidth]{M_example.png}
\end{center}
\end{figure}
\end{landscape}

# Supplemental Materials {-}
<!--
\setcounter{figure}{0}
\renewcommand\thefigure{S\arabic{figure}}
-->

\begin{landscape}
\begin{figure}
\caption{Median relative error of annual fully-selected fishing mortality rate in years 1 (Start), 21 (Middle), and 40 (End) from fitting simulated observations from each operating model with alternative process error assumptions in the estimating model. Estimating models also assume the mean natural mortality parameter $\beta_\text{M} = \log 0.2$ and no environmental covariate effect $\beta_\text{Ecov} = 0$.}\label{F_bias_M_fixed_beta_fixed}
\begin{center}
\includegraphics[height = \textheight]{F_bias_all_PE_effect_M_fixed_beta_fixed.png}
\end{center}
\end{figure}
\end{landscape}

\begin{landscape}
\begin{figure}
\caption{Median relative error of annual fully-selected fishing mortality rate in years 1 (Start), 21 (Middle), and 40 (End) from fitting simulated observations from each operating model with alternative process error assumptions in the estimating model. Estimating models also assume the mean natural mortality parameter $\beta_\text{M} = \log 0.2$ and the environmental covariate effect $\beta_\text{Ecov}$ is estimated.}\label{F_bias_M_fixed_beta_estimated}
\begin{center}
\includegraphics[height = \textheight]{F_bias_all_PE_effect_M_fixed_beta_estimated.png}
\end{center}
\end{figure}
\end{landscape}

\begin{landscape}
\begin{figure}
\caption{Median relative error of annual fully-selected fishing mortality rate in years 1 (Start), 21 (Middle), and 40 (End) from fitting simulated observations from each operating model with alternative process error assumptions in the estimating model. Estimating models also assume both the mean natural mortality parameter $\beta_\text{M}$ and the environmental covariate effect $\beta_\text{Ecov}$ are estimated.}\label{F_bias_M_estimated_beta_estimated}
\begin{center}
\includegraphics[height = \textheight]{F_bias_all_PE_effect_M_estimated_beta_estimated.png}
\end{center}
\end{figure}
\end{landscape}

### 2023-10-04 Ecov_study/mortality/paper/paper.tex
saving work on M-ecov working paper

@@

For the \(i\)th simulated data the relative error for a parameter
\(\theta\) provided from the fitted estimation model is \[
{\text{RE}_i}\left(\theta\right) = \frac{\widehat \theta_i - \theta_i}{\theta_i}
\] and measured bias as the median relative error and constructed 95\%
confidence intervals using the binomial distribution approach as in
\citet{stockmiller21}. If the confidence interval contains zero we do
not have evidence of bias from the simulation study. We used estimates
from all simulations that satisfied the first type of convergence
described above. We were not concerned about more restrictive types of
convergence because some estimation models would not be expected to
satisfy these criteria because of mis-specified structures. For example,
an estimation model that includes survival random effects fitted to a
set of observations simulated without survival random effects should
have trouble estimating variance of the survival random effects, but the
estimates of the parameters and derived output described above, might be
reliable.

@@

@@

@@

@@

>>>>

The parameters defining the simulated covariate time series, size of the
covariate effect, and any natural mortality random effects result in a
range of different levels of variation in annual natural mortality rates
(Figure \ref{M_example}).

@@

Results for fishing mortality rate are provided in the Supplementary
Materials. For the \(i\)th simulated data the relative error for a
parameter \(\theta\) provided from the fitted estimation model is \[
{\text{RE}_i}\left(\theta\right) = \frac{\widehat \theta_i - \theta_i}{\theta_i}
\] and measured bias as the median relative error and constructed 95\%
confidence intervals using the binomial distribution approach as in
\citet{stockmiller21}. If the confidence interval contains zero we do
not have evidence of bias from the simulation study. We used estimates
from all simulations that satisfied the first type of convergence
described above. We were not concerned about more restrictive types of
convergence because some estimation models would not be expected to
satisfy these criteria because of mis-specified structures. For example,
an estimation model that includes survival random effects fitted to a
set of observations simulated without survival random effects should
have trouble estimating variance of the survival random effects, but the
estimates of the parameters and derived output described above, might be
reliable.

For natural mortality rate, SSB, and fully-selected fishing mortality
rate, we summarized results for each of the annual values, but we
present results only for estimates from the first year (start), year 21
(middel), and the last year (end), because there were no appreciable
differences between the results for these three years and those for the
other years.

@@

We found high accuracy for estimation of the mean natural mortality rate
parameter (\(\beta_M\)) for all EM process errors assumptions when OMs
had step changes in fishing mortality, lower uncertainty in index and
age composition observations, and either R or R+M process errors (Figure
\ref{mean_M_bias}). The most variation in estimates occurred when
fishing and Higher uncertainty in index and age composition
observations. For OMs with R+S process errors, the most reliable
estimation of \(\beta_M\) was obtained when the EM also assumed R+S
process errors across all other factors. For OMs with R+M process
errors, the matching assumption for the EM only showed the best
reliability when there was low uncertainty in index and age composition
observations and fishing mortality was constant.

@@

We present results for error in annual natural mortality rate
conditional on three alternative EM configurations for the natural
mortality parameters: 1) mean natural mortality rate parameter is fixed
at the true value (\(\beta_M = log(0.2)\)) and no covariate effect is
assmed (\(\beta_\text{Ecov} = 0\)), 2) \(\beta_M = log(0.2)\) and
\(\beta_\text{Ecov}\) is estimated, and 3) both \(\beta_M\) and
\(\beta_\text{Ecov}\) are estimated.

When OMs and EMs assume \(\beta_M = log(0.2)\) and
\(\beta_\text{Ecov} = 0\), there is no annual variation in natural
mortality for OMs with R or R+S process errors simulated or assumed in
the EM and, therefore, estimation bias is not possible (Figure
\ref{annual_M_bias_M_fixed_beta_fixed}). We also observe little or no
evidence of bias (confidence intervals include 0) for R or R+S EMs for
any OMs even when the OMs included an effect of the covariate on natural
mortality (\(\beta_\text{Ecov} > 0\)). Including process error on M
(R+M) produces more variability in errors for R and R+S EMs than
including increased level of effect of the covariate on M. For R and R+S
OMs, errors in annual M for R+S EMs were less variable than those for R
EMs.

R+M EMs fit to R OMs with constant fishing mortality, or a step change
in fishing mortality and lower uncertainty in index and age composition
observations exhibited no bias in annual estimation of M indicating that
the variance of the estimated random effects for M in these EMs
collapsed to 0. R+M EMs were the only EMs to exhibit differences in the
sign of the median errors across the time series. When OMs had a step
change in fishing mortality and higher uncertainty in index and age
composition observations, we observed positive median errrors at the
beginning of the time series and lower median errors in later years.
However, whether there was evidence of bias, depended on the uncertainty
in covariate observations and the degree of variability in the latent
covariate. We observed negative median relative errors in annual M
estimates for R+M EMs fit to R+S OMs with lower uncertainty in index and
age composition observations across all levels of simulated effects of
the covariate, but evidence of bias was strongest at the beginning of
the time series.

When \(\beta_\text{Ecov}\) was estimated, R EMs performed worse for R+S
OMs with evidence of negative bias for OMs with cosntant fishing
mortality, lower uncertainty in index and age composition observations,
the largest covariate effect size, lower variability in the latent
covariate, and larger uncertainty in the covariate observations (Figure
\ref{annual_M_bias_M_fixed_beta_estimated}). There was little difference
in the results for R+S EMs whether \(\beta_\text{Ecov}\) was estimated
or not. The results for R+M EMs were generally similar to those when
\(\beta_\text{Ecov}=0\) was assumed, except that the differences between
the estimates at the beginning of the time series and later years for
certain OM configurations did not occur.

Allowing the EMs to estimate \(\beta_M\) resulted in very large
variability in estimates for all EMs for R and R+M OMs with constant
fishing mortality and higher uncertainty in index and age composition
observations (Figure \ref{annual_M_bias_M_estimated_beta_estimated}).
The same varibilities occurred for R+S OMs, but strong bias was
estimated for R and R+M EMs. Annual M estimation was most reliable when
OMs had step changes in fishing mortality and lower uncertainty in index
and age composition observations. For R and R+M OMs with those
characteristics, all EMs generally provided accurate estimation of
annual M, but only R+S EMs provided accurate estimation for R+S OMs.
Little or no bias in annual M estimation was observed for R+S EMs across
all OM process error assumptions as long as there was a step change in
fishing mortality and lower uncertainty in indices and age composition.

@@

Like natural mortality, we present results for error in spawning biomass
conditional on three alternative EM configurations for the natural
mortality parameters: 1) mean natural mortality rate parameter is fixed
at the true value (\(\beta_M = log(0.2)\)) and no covariate effect is
assmed (\(\beta_\text{Ecov} = 0\)), 2) \(\beta_M = log(0.2)\) and
\(\beta_\text{Ecov}\) is estimated, and 3) both \(\beta_M\) and
\(\beta_\text{Ecov}\) are estimated.

When EMs assume \(\beta_M = log(0.2)\) and \(\beta_\text{Ecov} = 0\),
there is little evidence of bias for R and R+M OMs except when there is
a step change in fishing mortality and higher uncertainty in index and
age composition observations, particularly at the end of the time series
(Figure \ref{SSB_bias_M_fixed_beta_fixed}). We observed a similar trend
in median relative error for R+S OMs, but there was more evidence of
positive bias at the beginning of the time series whereas the confidence
intervals for the negative median errors at the end of the time series
often included 0. However, there was more indication of positive bias of
the incorrect process error assumption of the EMs for the R+S OMs. For
R+S EMs fit to R+S OMs, there was indication of small positive bias at
the beginning of the time series when there was constant fishing
mortality and higher uncertainty in index and age composition
observations.

When EMs assume \(\beta_M = log(0.2)\), but estimate
\(\beta_\text{Ecov}\), the median errors for SSB for R and and R+M OMs
are similar to those when \(\beta_\text{Ecov} = 0\) (Figure
\ref{SSB_bias_M_fixed_beta_estimated}). For R+S OMs, the median errors
for EMs that also assume R+S process errors are also similar to those
with no covariate effect assumed. However, for R and R+M EMs fit to R+S
OMs with constant fishing mortality rate and lower uncertainty in
indices and age composition observations, we observed evidences of
negative bias when higher covariate observation uncertainty and lower
variation in the latent covariate was simulated and positive bias for
other configurations of the covariate and corresponding observation
uncertainty. When R+S OMs had constant fishing mortality rate and higher
uncertainty in indices and age composition observations or step changes
in fishing mortality and lower uncertainty in indices and age
composition, R and R+M EMs often provided positively biased SSB
estimates when there was lower uncertainty in covariate observations.

When EMs estimated both \(\beta_M\) and \(\beta_\text{Ecov}\), we
observed large variation in the errors in SSB under the same OM
configurations where we observed large variation in errors for annual
natural mortality rates (Figure
\ref{SSB_bias_M_estimated_beta_estimated}). Similarly, we observed
little or no bias in SSB estimation for R+S EMs across all OM process
error assumptions as long as there was a step change in fishing
mortality and lower uncertainty in indices and age composition.

@@

\begin{landscape}
\begin{figure}
\caption{Example simulations of annual natural mortality rates that may be a function of a temporally varying environmental covariate and autoregressive random effects.}\label{M_example}
\begin{center}
\includegraphics[width = \textwidth]{M_example.png}
\end{center}
\end{figure}
\end{landscape}

\hypertarget{supplemental-materials}{%
\section*{Supplemental Materials}\label{supplemental-materials}}
\addcontentsline{toc}{section}{Supplemental Materials}

\begin{landscape}
\begin{figure}
\caption{Median relative error of annual fully-selected fishing mortality rate in years 1 (Start), 21 (Middle), and 40 (End) from fitting simulated observations from each operating model with alternative process error assumptions in the estimating model. Estimating models also assume the mean natural mortality parameter $\beta_\text{M} = \log 0.2$ and no environmental covariate effect $\beta_\text{Ecov} = 0$.}\label{F_bias_M_fixed_beta_fixed}
\begin{center}
\includegraphics[height = \textheight]{F_bias_all_PE_effect_M_fixed_beta_fixed.png}
\end{center}
\end{figure}
\end{landscape}

\begin{landscape}
\begin{figure}
\caption{Median relative error of annual fully-selected fishing mortality rate in years 1 (Start), 21 (Middle), and 40 (End) from fitting simulated observations from each operating model with alternative process error assumptions in the estimating model. Estimating models also assume the mean natural mortality parameter $\beta_\text{M} = \log 0.2$ and the environmental covariate effect $\beta_\text{Ecov}$ is estimated.}\label{F_bias_M_fixed_beta_estimated}
\begin{center}
\includegraphics[height = \textheight]{F_bias_all_PE_effect_M_fixed_beta_estimated.png}
\end{center}
\end{figure}
\end{landscape}

\begin{landscape}
\begin{figure}
\caption{Median relative error of annual fully-selected fishing mortality rate in years 1 (Start), 21 (Middle), and 40 (End) from fitting simulated observations from each operating model with alternative process error assumptions in the estimating model. Estimating models also assume both the mean natural mortality parameter $\beta_\text{M}$ and the environmental covariate effect $\beta_\text{Ecov}$ are estimated.}\label{F_bias_M_estimated_beta_estimated}
\begin{center}
\includegraphics[height = \textheight]{F_bias_all_PE_effect_M_estimated_beta_estimated.png}
\end{center}
\end{figure}
\end{landscape}
